### 2014-08-13 doc/thesis/typedlua.tex
Rewriting the section about table types and interfaces

Typed Lua includes table types to represent Lua's tables.
Although Typed Lua has just one table type constructor, it has three
different syntaxes for defining the type of a table.
The first syntax defines table types for hash tables;
it is written \texttt{\{t1:t2\}},
and maps to the table type $\{t_{1}:t_{2}\;\cup\;\Nil\}$.
This table type represents a hash table that maps values of type
$t_{1}$ to values of type $t_{2}\;\cup\;\Nil$, as Lua returns
$\Nil$ when we use a non-existing key to index a table.
The second syntax defines table types for lists;
it is written \texttt{\{t\}},
and maps to the table type $\{\Number:t\;\cup\;\Nil\}$.
This table type is just a syntactic sugar to the definition of a
hash table that maps values of type $\Number$ to values of type
$t\;\cup\;\Nil$.
The third syntax defines table types for records;
it is written \texttt{\{s1:t1, ..., sn:tn\}},
and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$.
This table type represents a record that maps the literals
$s_{i}$\texttt{, ..., }$s_{n}$ to values of types
$t_{i}$\texttt{, ..., }$t_{n}$, where $i$ may vary from one to $n$.

Next, we show one example of table type to type a hash table:
\begin{verbatim}
    local t:{string:number} = { foo = 1 }  -- OK
    local x:number = t.foo                 -- Error
    local y:number? = t.bar                -- OK
\end{verbatim}

The second line of this example gives a type error for the same
reason that the last line of the previous example gives a type error
too, that is, we are trying to assign a value that can be of two
different types to a variable that accepts values of only one type.
Although the field \texttt{bar} does not exist in \texttt{t}, the last
line of this example does not give a type error because the
annotated type matches the type of the values that can be stored in
\texttt{t}.

Next, we show one example of table type to type a list:
\begin{verbatim}
    local days:{string} = { "Sunday", "Monday", "Tuesday",
                            "Wednesday", "Thursday", "Friday",
                            "Saturday" }  -- OK
\end{verbatim}

When we know that a list has fixed elements, we can leave the
variable declaration unannotated and let local type inference assign
a more specific table type to the list.
If we remove the annotation in this example, the compiler uses the
syntax of records to infer the following table type to \texttt{days}:
\begin{align*}
\{{1:\String},\;{2:\String},\;{3:\String},\;{4:\String},\;\\
{5:\String},\;{6:\String},\;{7:\String}\}
\end{align*}
This inferred table type is a subtype of the type
$\{\Number:\String\;\cup\;\Nil\}$, which is the type that we used to
annotate the variable \texttt{days}.

@@

It also worth mentioning that we can use the style of the inferred
type to define heterogeneous tuples.

@@

Typed Lua includes interfaces as syntactic sugar to table types
that specify records.
For instance, we can use the following interface as an alias to the
type of the record we showed in the previous example:
\begin{verbatim}
    interface Person
      firstname:string
      lastname:string
    end
\end{verbatim}

Now we can use the type \texttt{Person} as an alias to a table type
that certainly has the fields \texttt{firstname} and \texttt{lastname}.
Next, we show an example of code that uses the type \texttt{Person}:
\begin{verbatim}
    local function greeter (person:Person):string
      return "Hello " .. person.firstname ..
             " " .. person.lastname
    end

@@

    print(greeter(user1)) -- OK
    print(greeter(user2)) -- Error
    print(greeter(user3)) -- OK
    print(greeter(user4)) -- Error
\end{verbatim}

@@

Using interfaces we can also define recursive types such as linked
lists:

\begin{verbatim}
    interface Element
      info:number
      next:Element?
    end
\end{verbatim}

We need to explicitly annotate \texttt{next} being of type
\texttt{Element}$\;\cup\;\Nil$ because in our type system $\Nil$
is not the bottom type, that is, $\Nil$ is subtype of itself only.
Typed Lua has this behavior because in Lua $\Nil$ represents
the absence of an useful value.

@@

Typed Lua includes function types because functions are first-class
values in Lua.
The syntax for defining a function type is \texttt{(s) -> (s)},
where \texttt{s} is a second-class type.
As we saw in Figure \ref{fig:typelang}, a second-class type in
Typed Lua can be the type $\Void$, a first-class type, a variadic
type, or a tuple type.
We use tuple types to represent multiple parameters, multiple return
values, and multiple assignments, and we use the variadic type to
represent variadic functions and the vararg expression.
Even though we already included the tuple type and the vararg type,
we are still working on these types.

The next example shows how we can annotate a \texttt{map} function
that receives another function as its parameter and uses this
function to alter the elements of a list of numbers, which is
also a parameter of \texttt{map}:
\begin{verbatim}
    local function map (f:(number)->(number), l:{number}):void
      for i=1, #l do
        l[i] = f(l[i] or 0)
      end
    end
\end{verbatim}

To conclude this section, we just started the design and
implementation of Typed Lua, and there is still several design
decisions to be taken on this subject.
For instance, our type system should handle the subtleties that
come with metatables.

@@

>>>>

Tables are the only mechanism that Lua has to build data structures;
they are associative arrays where any value (except \texttt{nil})
can be used as a key.
Programmers can use tables to represent tuples, arrays (dense or sparse),
records, graphs, modules, objects, etc.
Lua has syntactic sugar for indexing tables as records:
\texttt{t.k} is syntactic sugar for \texttt{t["k"]}, and means that
we are accessing field \texttt{k} from table \texttt{t}. 
In this section, we show how Typed Lua encodes tables as maps,
arrays, and records.

Table types are the only framework that Typed Lua has to represent
the different uses that a Lua table has.
A table type $\{K_{1}:T_{1}, ..., K_{n}:T_{n}\}$ represents a map
from values of type $K_{i}$ to values of type $T_{i}$.

Although Typed Lua has just one framework to represent tables,
it has three different syntaxes for defining the type of a table.
The first syntax defines table types for maps:
it is written \texttt{\{k:t\}},
and maps to the table type $\{K:T\;\cup\;\Nil\}$.
This table type represents a map that maps values of type
$K$ to values of type $T\;\cup\;\Nil$,
as Lua returns $\Nil$ when we use a non-existing key to index a table.
The second syntax defines table types for arrays:
it is written \texttt{\{t\}},
and maps to the table type $\{\Number:T\;\cup\;\Nil\}$.
This table type is just a syntactic sugar to the definition of a
map that maps values of type $\Number$ to values of type $T\;\cup\;\Nil$.
The third syntax defines table types for records:
it is written \texttt{\{k1:t1, ..., kn:tn\}},
and maps to the table type $\{K_{1}:T_{1}, ..., K_{n}:T_{n}\}$.
This table type represents a record that maps the literal types
$K_{i}$\texttt{, ..., }$K_{n}$ to values of types
$T_{i}$\texttt{, ..., }$T_{n}$, where $i$ may vary from one to $n$.

Next, we show one example of table type to type a map from strings to numbers:
\begin{verbatim}
    local t:{string:number} = { foo = 1 } 
    local x:number = t.foo               -- compile-time error
    local y:number? = t.bar              -- y gets nil
    local z:number = t["bar"] or 0       -- z gets 0 
\end{verbatim}

The second line of this example gives a type error because we are
attempting to assign a value of type $\Number\;\cup\;\Nil$ to a
variable that accepts only values of type $\Number$.
Although the field \texttt{bar} does not exist in \texttt{t}, the third
line of this example does not give a type error because the
annotated type matches the type of the values that can be stored in
\texttt{t}.
The last line shows that the \texttt{or} idiom is also useful to
give a default value to a missing table field.

Next, we show one example of table type to type an array:
\begin{verbatim}
    local days:{string} = { "Sunday", "Monday", "Tuesday",
                            "Wednesday", "Thursday", "Friday",
                            "Saturday" }
    local x = days[1]                    -- x gets "Sunday"
    local y = days[8]                    -- y gets nil
\end{verbatim}

In this example, the type of both variables \texttt{x} and \texttt{y}
is $\String\;\cup\;\Nil$.
An incovenience aspect of making the types of maps and arrays always
include the type $\Nil$ is to overload the programmers,
as they need to use \texttt{or} or \texttt{if} to narrow the type of
the elements they are accessing.

When we know that a list has fixed elements, we can leave the
variable declaration unannotated and let local type inference assign
a more specific table type to the variable.
If we remove the annotation in this example, the compiler uses the
syntax of records to infer the following table type to \texttt{days}:
\begin{align*}
\{{1:\String},\;{2:\String},\;{3:\String},\;{4:\String},\;\\
{5:\String},\;{6:\String},\;{7:\String}\}
\end{align*}
% text about subtyping

@@

We can also use the style of the inferred type to define heterogeneous
tuples:
\begin{verbatim}
    local album = { "Transformer", 1972, "Lou Reed" }
\end{verbatim}

The next example shows how we can annotate a \texttt{map} function
that receives another function as its parameter and uses this
function to alter the elements of a list of numbers, which is
also a parameter of \texttt{map}:
\begin{verbatim}
    local function map (f:(number)->(number), l:{number})
      for i = 1, #l do
        l[i] = f(l[i] or 0)
      end
    end
\end{verbatim}

@@

Typed Lua includes \emph{interfaces} as syntactic sugar to table types
that specify records, as writing table types can be unwieldy when
records get bigger and types of record fields get more complicated.
For instance, we can use the following interface as an alias to the
type of the record we showed in the previous example:
\begin{verbatim}
    local interface Person
      firstname:string
      lastname:string
    end
\end{verbatim}

The declaration of the interface \texttt{Person} is an alias to the table type
$\{``firstname":\String, ``lastname":\String\}$ in the remainder
of the current scope.
Now we can use \texttt{Person} in type declarations:
\begin{verbatim}
    local function greeter (person:Person):string
      return "Hello " .. person.firstname ..
             " " .. person.lastname
    end

@@

    print(greeter(user1)) -- Hello, Lewis Allan
    print(greeter(user2)) -- compile-time error
    print(greeter(user3)) -- Hello, Lou Reed
    print(greeter(user4)) -- compile-time error
\end{verbatim}

@@

If our record type has optinal fields, that is, fields that can be \texttt{nil},
we need to use an explicit type declaration when declaring a
variable of this record type, as we show in the following example:
\begin{verbatim}
    local interface Person
      firstname:string
      middlename:string?
      lastname:string
    end
    local user1:Person = { firstname = "Lewis",
                           middlename = "Allan" or nil,
                           lastname = "Reed" }
    local user2:Person = { lastname = "Reed",
                           firstname = "Lou" }
\end{verbatim}

We need an explicit type declaration because neither of the types
that the Typed Lua compiler assigns to the above table constructors
is a subtype of
$\{``firstname":\String, ``middlename":\String\cup\Nil, ``lastname":\String\}$,
the type that \texttt{Person} describes.
While declaring \texttt{user1}, we use \texttt{or} to cast the type of
\texttt{middlename} because of the contravariance in the type of
mutable fields.

We can also use interfaces to define recursive types such as liked lists:

\begin{verbatim}
    local interface Element
      info:number
      next:Element?
    end
\end{verbatim}

The declaration of the interface \texttt{Element} is an alias to the
recursive type
$\mu Element.\{``info":\Number, ``next":Element \cup \Nil\}$ in the
remainder of the current scope.

We need to explicitly annotate \texttt{next} being of type
$Element\;\cup\;\Nil$ because in our type system $\Nil$
is not the bottom type, that is, $\Nil$ is subtype of itself only.
Typed Lua has this behavior because in Lua $\Nil$ represents
the absence of an useful value.
Thus, making $\Nil$ the bottom type would severaly reduce the amount
of type safety in the system.

Now we can use \texttt{Element} to annotate a function that
inserts an element at the begining of a list:
\begin{verbatim}
    local function insert (e:Element?, v:number):Element 
      return { info = v, next = e }
    end
\end{verbatim}

We need an explicit type declaration on the return type because
the Typed Lua compiler cannot infer recursive types, though
it has subtyping rules to check that the inferred return type
matches the annotation.

Now we can write a program that declares a list, inserts some
elements in this list, and then traverses the list printing
each element of the list:
\begin{verbatim}
    local l:Element?

    l = insert(l, 4)
    l = insert(l, 3)
    l = insert(l, 2)
    l = insert(l, 1)

    while l do
      if l then print(l.info) end
      l = l.next                  -- compile-time error
    end
\end{verbatim}

However, this example has one compile-time error because we
are attempting to index the variable \texttt{l},
which can be \texttt{nil}.
Notice that we would have the same compile-time error while
printing the list element if we had not narrowed the type of
\texttt{l}.
To eliminate this compile-time error we have to write a
function that narrows the type of \texttt{l} before using it:
\begin{verbatim}
    local function get_next (l:Element?)
      if l then return l.next else return nil end
    end
\end{verbatim}

Now we can use \texttt{get\string_next} to rewrite the \texttt{while}
from the previous example, thus eliminating the compile-time error:
\begin{verbatim}
    while l do
      if l then print(l.info) end
      l = get_next(l)
    end
\end{verbatim}

@@

@@

Rule for checking if a table type is well formed:
\[
\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
\]
